How to improve your relationship with your future self.

### 2016-06-05 workflow.Rmd
Workflow piece almost ready for Maarten to edit

---
title: How to improve your relationship with your future self.
author: 
- name: Jake Bowers\thanks{I owe many thanks to Mark Fredrickson,
    Brian Gaines, Kieran Healy, Kevin Quinn and Cara
    Wong for direct help on this document and to Mika LaVaque-Manty
    and Ben Hansen for many useful discussions on this topic. The
    source code for this document may be freely downloaded and
    modified from \url{https://github.com/jwbowers/workflow}.}
  affiliation: <jwbowers@illinois.edu>, University of Illinois at Urbana-Champaign and White House Social and Behavioral Sciences Team
- name: Maarten Voors
  affiliation: Wageningen University
bibliography: ../BIB/references.bib
csl: american-political-science-review.csl
date: 19 May 2016
...

@@

## To make the html file do
## render("exploration4.Rmd",output_format=html_document(fig_retina=FALSE))
## To make the pdf file do
## render("exploration4.Rmd",output_format=pdf_document())

@@

options(SweaveHooks=list(fig=function(){
			   par(mar=c(3.5, 3, 1.1, 0),
			       pty="s",
			       mgp=c(1.5,0.5,0),
			       oma=c(0,0,0,0))},
			 echo=function(){options(continue=" ") ##Don't show "+" prompts,
			 options(prompt=" ")
			 }),
	digits=4,
	scipen=8,
	width=132
	)
options(error=function(){options(prompt="> ",continue="+ ");NULL})
```

| Do I contradict myself?
| Very well then I contradict myself,
| (I am large, I contain multitudes.)
| [@whitman1855song]

> If you tell the truth, you don't have to remember anything. [@twain1976book]^[<https://www.marktwainhouse.org/man/famous_twain_quotes.php>]

> Never memorize what you can look up in books -Albert Einstein^[<https://en.wikiquote.org/wiki/Albert_Einstein>]

Memory is tricky. When we really want to remember something we should practice it until it becomes internalized. In fact, learning requires effort. When we do not practice, we tend to be overconfident in our recall abilities [@koriat2005illusions] and false memories can be implanted.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.]

@@

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised version of @bowers2011tpm and provides some suggestions for practices that will make such reproduction occur much more easily and quickly in the event that famous papers require special scrutiny.  Specifically, this piece aims to amplify some of what we already ought to know,  and to add to some of those ideas
given current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with teh DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese1007replication and @asendorpf2013recommendations.] This essay is an attempt at providing practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on other bad past decisions but not on decisions about data analysis and the production of scholarly papers.

%```{r read.data,echo=FALSE,results=hide}
## This data frame will be used throughout the analyses.
## Putting it up top, so we can use it in the comments section.
##  This read.csv line is slightly more complex than needed because of
##  the particular encoding used for characters in the source file.
dem.nations.df<-read.csv(url("http://www.hks.harvard.edu/fs/pnorris/Data/Democracy%20CrossNational%20Data/Democracy%20Crossnational%20Data%20Spring%202009.csv",encoding="ISO_8859-1"),as.is=TRUE)
row.names(dem.nations.df)<-dem.nations.df$Natabrv ##make the row names more helpful
good.df<-na.omit(dem.nations.df[,c("Gini2004","protac2000","Nation","meanpr")])
good.df$gini04<-good.df$Gini2004/100
good.df$protac00<-good.df$protac2000
```

\section{Data analysis is computer programming.}\label{sec:data-analys-comp}

@@

Using R \citep{R:2011}, for example, I might specify that the file
\Verb+fig1.pdf+ was produced by the following commands in a file
called \Verb+fig1.R+.\footnote{The command \Verb+please-plot+ and some
  other R functions used in this essay come from the
  \Verb+MayIPleaseDoStatistics+ package which emphasizes politeness in
  data analysis. Functions like \Verb+please-plot+ can be blocked and
  more polite versions such as \Verb+may-I-please-have-a-plot+ can be
  required using \Verb+options(politeness=99)+ }

\begin{Verbatim}[fontsize=\footnotesize]
thedata <- read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf('fig1.pdf') ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2<-please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
\end{Verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was
created, I will know: (1) ``that plot'' is from a file called
\Verb+fig1.pdf+ and (2) \Verb+fig1.pdf+ was created in
\Verb+fig1.R+. In a future where R still exists, changing the figure
will require quick edits of commands already written.  In a future
where R does not exist, I will at least be able to read the plain text
R commands and use them to write code in my new favorite statistical
computing language: R scripts are written in
\href{http://en.wikipedia.org/wiki/Plain_text}{plain text}, and plain
text is a format that will be around as long as computer programmers
write computer programs.\footnote{Since R is open
  source, I will also be able to download an old version of R, download an
  old-fashioned open-source operating system (like Ubuntu 10), and
  run the old-fashioned statistical computing environment in the
  old-fashioned operating system in a virtual machine on my
  new-fashioned actual machine.}

Moreover, realize that file names send messages to your future
self. Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named \Verb+inequality-and-protest-figures.R+ than
if your files are called \Verb+temp9+ or
\Verb+supercalifragilisticexpialidocious+. The extension
\Verb+.R+ tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our
antique hard drives for files containing R scripts.

\paragraph{Step \thesection} If we know the provenance of results, 
future  or current collaborators can quickly and easily reproduce
and thus change and improve upon the work.

\section{No data analyst is an island for long.}
Data analysis involves a long series of decisions. Each decision
requires justification.  Some decisions will be too small and
technical for inclusion in the published article itself. These need to
be documented in the code itself \citep{nagler1995coding}. Paragraphs
and citations in the publication will justify the most important
decisions. So, one must code to communicate with yourself and
others. There are two main ways to avoid forgetting the reasons you
did something with data: comment your code and tightly link your code
with your writing.\footnote{One can also try the R command
  \Verb+put-it-in-my-mind(reason,importance='high')+ to firmly place a
  reason for a decision into the mind of the analyst. I myself have
  not had much luck with this function.}

\subsection{Code to communicate: Comment your code.}
Comments --- unexecuted text inside of a script --- are a message to
collaborators (including your future self) and other consumers of your
work. In the above code chunk, I used comments to explain the lines to
readers unfamiliar with R and to remember that I had tried a different
specification but decided not to use it because adding the squared
term did not really change the substantive story arising from the
model.\footnote{R considers text marked with \texttt{\#} as a comment.}
Messages left for your future self (or near-future others) help
retrace and justify your decisions as the work moves from seminar
paper to conference paper to poster back to paper to dissertation and
onwards.

% Luckily, if you are
% using a literate programming practice (ex. Sweave or odfWeave), you
% can write paragraphs to surround your code as well as technical
% comments in the code itself. If you are not using a strictly literate
% programming practice then use whatever commenting protocol exists in
% your analysis language. R, for example, considers text marked with
% ``\#'' as a comment. Here for example I comment about a linear model
% run for a paper and also include comments about another model run for
% due diligence.

% ```{r reg1,echo=TRUE,results=verb}
% ##  Repeat the regression run for Table~\ref{tab:protest}
% ##  Do political rights predict protest reporting?
% lm1 <- lm(protac00 ~ gini04 + meanpr, data=good.df)
% ##  Q: Was it worthwhile controlling for political rights?
% lm2 <- lm(protac00~gini04,data=good.df)
% coef(lm2)[2]-coef(lm1)[2]
% ##  A: Yes (around 3/4 an act of protest difference)
% ```

@@

Comment liberally. Comments are discarded when R runs analysis or
\LaTeX~turns plain text into page images, so only those who dig into
your work will see them.

\subsection{Code to communicate: Literate programming.}

\begin{quote}
  Let us change our traditional attitude to the construction of
  programs: Instead of imagining that our main task is to instruct a
  computer what to do, let us concentrate rather on explaining to
  human beings what we want a computer to do.
  \citep[][p. 97]{knuth1984literate}
\end{quote}

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf file or some other format which displays page
images to ease reading rather than to encourage reanalysis and
rewriting. Eventually pieces of that report (tables, graphs,
paragraphs) ought to show up in, or at least inform, the publishable
paper. Re-creating those analyses by pointing, clicking, copying, or
pasting would invite typing error and waste time. Re-creating your
arguments justifying your analysis decisions would also waste
time. More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the ``source code'' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.

How might one avoid these problems?
\href{http://en.wikipedia.org/wiki/Literate_programming}{Literate
  programming} is the practice of weaving code into a document ---
paragraphs, equations, and diagrams can explain the code, and the code
can produce numbers, figures, and tables (and diagrams and even equations and
paragraphs). Literate programming is not merely fancy commenting but
is about enabling the practice of programming itself to facilitate easy
reproduction and communication.

For example, in \S\ref{sec:data-analys-comp}, I suggested that we know
where ``that plot on page 10'' comes from by making sure we had a
\Verb+fig1.pdf+ file produced from a clearly commented plain text file
called something like \Verb+fig1.R+.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
\LaTeX~ markup with R code chunks to make things like
Figure~\ref{fig:giniprot}. This combination of \LaTeX and R is called
Sweave \citep{Leis:2005}.\footnote{Support for Sweave is included with
  R.}

\begin{Verbatim}[fontsize=\footnotesize]
This paper, for example, was written in plain text using \LaTeX
markup with R code chunks to make things like
Figure~\ref{fig:giniprot}. This combination of \LaTeX and R is
called Sweave \citep{Leis:2005}.\footnote{Support for Sweave is
  included with R.}
\begin{figure}[h]
\centering
 ```{r fig1plot,fig=TRUE}
 ##  Make a scatterplot of Protest by Inequality
 with(good.df,plot(gini04,protac00,xlab='Gini Coefficient 2004 (UNDP)',
                   ylab='Mean Protest Activities\n(World Values Survey 1980-2000)'))
 ##  Label a few interesting points
 with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
      text(gini04,protac00,labels=Nation))
 ```
\caption{Protest activity by income inequality \citep[from][]{norris2009data}.}
\label{fig:giniprot}
\end{figure}
\end{Verbatim}

```{r fig1code,echo=FALSE,fig=FALSE}
##  Make a scatterplot of Protest by Inequality
with(good.df,plot(gini04,protac00,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
                  cex=.8))

##  Label a few interesting points
with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(gini04,protac00,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
```

\begin{figure}[H]
  \begin{center}
```{r fig1plot,echo=FALSE,fig=TRUE,width=4,height=4}
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
```{r fig1code>>
```
    \caption{Protest activity by income inequality \citep[from][]{norris2009data}.}
    \label{fig:giniprot}
  \end{center}
\end{figure}

By using \Verb+\label{fig:giniprot}+, I do not need to keep track of
the figure number, nor do extra work when I reorganize the document in
response to reviewer suggestions. Nor do I need a separate
\Verb+fig1.R+ file or \Verb+fig1.pdf+ file.  Tables and other
numerical results are also possible to generate within the source
code of a scholarly paper. Those who view the code for this essay will
see how Table~\ref{tab:protest} was also generated directly from a
regression object.\footnote{\citet{beck2010reg} inspired this
  particular presentation of a linear model.}

```{r maketable, echo=FALSE, results=hide}
  lm1<-lm(protac00~gini04+meanpr,data=good.df)  #  Run the regression

@@

library(xtable)
thextab<-xtable(themat,align=c("l",rep(".",4)),digits=1,
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from \\citep{norris2009data}.",label="tab:protest")
```

```{r printtable,echo=FALSE,results=tex}
  print(thextab,include.colnames=FALSE,
        hline.after=NULL,
        table.placement="H",
        add.to.row=list(
          pos=list(0,3),
          command=c(
            "\\toprule \n &
             \\multicolumn{1}{c}{Coef} &
             \\multicolumn{1}{c}{Std. Err.} &
             \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
            paste("& \\multicolumn{4}{c}{n: ",n,",",
                  "   resid.sd: ",signif(est.sigma),
                  ", R$^2$: ",r2,"} \\\\ \\bottomrule",sep="")
            )
          ))
```

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.\footnote{The R project has a task view devoted to
  \href{http://cran.r-project.org/web/views/ReproducibleResearch.html}{reproducible
    research} listing many of the different approaches to literate
  programming for R.}  \href{http://www.lyx.org/}{LyX} offers a
WYSIWYG environment for \LaTeX~ that supports Sweave. And the
\Verb+odfWeave+ package in R allows the use of OpenOffice documents in
exactly the same way.\footnote{A quick Google search of ``Sweave for
  Stata'' turned up lots of resources for literate programming with
  Stata.} If your workflow does not involve \LaTeX~ and R, you can
still implement some of the principles here. Imagine creating a style
in Microsoft Word called ``code'' which hides your code when you print
your document, but which allows you to at least run each code chunk
piece by piece [or perhaps there are ways to extract all text of style
``code'' from a Microsoft Word document]. Or imagine just using some
other kind of indication linking paragraphs to specific places in code
files. There are many ways that creative people can program in a
literate way.

Literate programming need not go against the principle of modular data
analysis \citep{nagler1995coding}. In my own work I routinely have
several different Sweave files that fulfill different functions, some
of them create \LaTeX code that I can \Verb+\input+ into my
\Verb+main.tex+ file, others setup the data, run simulations, or allow
me to record my journeys down blind alleys. Of course, when we have
flying cars running on autopilot, perhaps something other than
Sweave will make our lives even easier. Then we'll change.

\paragraph{Step \thesection}

We analyze data in order to explain something about the world to other
scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis to human beings, we will do a better job
with the data analysis itself: we will learn as we focus on teaching,
and we will avoid errors and save time as we ensure that others
(including our future selves) can retrace our steps. A document than
can be ``run'' to reproduce all of the analyses also instills
confidence in readers and can more effectively spur discussion and
learning and cumulation of research.

\section{Meaningful code requires data.}

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on (``load''/ ``open'' / ``get'' /
``use'') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
\Verb+load('thedata.rda')+ or \Verb+read.csv('thedata.csv')+ or some
equivalent line in them, and \Verb+thedata.csv+ should be stored in
some place easy to find (like in the same directory as the file or
perhaps in \Verb+'Data/thedata.rda'+). Of course, it never hurts to
drop in a comment pointing to the data file.

Where should one store data files? An obvious solution is always to
make sure that the data file used by a command file is in the same
directory as the command file. More elegant solutions require all
co-authors to have the same directory structure so that
\Verb+load('Data/thedata.rda')+ means the same thing on all computers
used to work on the project. This kind of solution is one of the
things that formal version control systems do well (as discussed
a bit more in \S\ref{sec:vers-contr-prev}).

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
\citep{nagler1995coding}. So, perhaps your analysis oriented files
will \Verb+load('cleandata.rda')+ and a comment in the code will alert
the future you (among others) that \Verb+cleandata.rda+ was created
from \Verb+create-cleandata.R+ which in turn begins with
\Verb+read.csv(\url('http://www.greatfreedata.gov/dirtydata.csv'))+. Such
a data processing file will typically end with something like
\Verb+save('cleandata.rda')+ so that we are doubly certain about the
provenance of the data.\footnote{Of course, if you need math or
  paragraphs to explain what is happening in these files, you might
  prefer to make them into Sweave files, for which the conventional
  extension is .Rnw. So you'd have \Verb+create-cleandata.Rnw+ which
  might explain and explore the different coding decisions you made,
  perhaps involving a factor analysis and diagnostic plots.}

Now, if in the future we wonder where \Verb+cleandata.rda+ came from, we
might search for occurrences of `cleandata' in the files on our
system. However, if such searching among files is a burden, an even
nicer solution is to maintain a file for each project called
``MANIFEST.txt'' or ``INDEX.txt'' or ``README.txt'' which lists the
data and command files with brief descriptions of their functions and
relations.

\paragraph{Step 3} We should know where the data came
from and what operations were performed on which set of data.

@@

\begin{Verbatim}[fontsize=\footnotesize]
DA NI=19 NO=199 MA=CM
LA=basic.lab
CM FI=basic.cov
\end{Verbatim}

>>>>

---
title: How to improve your relationship with your future self.
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>, University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team. Many thanks to Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn and Cara Wong for direct help on this document and to Mika LaVaque-Manty and Ben Hansen for many useful discussions on this topic. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "--filter", "/usr/local/bin/pandoc-crossref"
    ]
...

@@

## To make the html file do
## render("exploration4.Rmd",output_format=html_document(fig_retina=FALSE))
## To make the pdf file do
## render("exploration4.Rmd",output_format=pdf_document())
## to debug, try render("thefile.Rmd", clean=FALSE) so that you can see the intermediate files

@@

options(SweaveHooks=list(fig=function(){
			   par(mar=c(3.5, 3, 1.1, 0),
			       pty="s",
			       mgp=c(1.5,0.5,0),
			       oma=c(0,0,0,0))},
			 echo=function(){options(continue=" ") ##Don't show "+" prompts,
			 options(prompt=" ")
			 }),
	digits=4,
	scipen=8,
	width=132
	)

#options(error=function(){options(prompt="> ",continue="+ ");NULL})
```

> | Do I contradict myself?
| Very well then I contradict myself,
| (I am large, I contain multitudes.)
| [@whitman1855song]

> If you tell the truth, you don't have to remember anything. [@twain1975mark]

Memory is tricky. When we really want to remember something we should practice it until it becomes internalized. In fact, learning requires effort. When we do not practice, we tend to be overconfident in our recall abilities [@koriat2005illusions] and false memories can be implanted.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] If we cannot count on memory, then how can we do science?

@@

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised version of @bowers2011tpm and provides some suggestions for practices that will make such reproduction occur much more easily and quickly in the event that famous papers require special scrutiny.  Specifically, this piece aims to amplify some of what we already ought to know,  and to add to some of those ideas
given current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations.] This essay is an attempt at providing practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on other bad past decisions but not on decisions about data analysis and the production of scholarly papers.

```{r read.data,echo=FALSE,results='hide', cache=TRUE}
## This data frame will be used throughout the analyses.
## Putting it up top, so we can use it in the comments section.
library(readstata13)
dem.nations.df<-read.dta13("https://www.dropbox.com/s/huhqag5rudwtbno/Democracy%20Cross-National%20Data%20V4.1%2009092015.dta?dl=1&pv=1&raw=1",convert.factors=FALSE)

row.names(dem.nations.df)<-dem.nations.df$Natabrv ##make the row names more helpful
good.df<-na.omit(dem.nations.df[,c("Gini2004","WVS5_protest","Nation","meanpr")])
good.df$gini04<-good.df$Gini2004/100
good.df$protest05<-good.df$WVS5_protest
```

We organize the paper around a series of homilies that lead to certain concrete actions.

# Data analysis is computer programming. {#sec:data-analys-comp}

@@

Using R [@R:2011], for example, I might specify that the file `fig1.pdf` was
produced by the following commands in a file called `fig1.R`.^[The command
`please-plot`and some other R functions used in this essay come from the
`MayIPleaseDoStatistics` package which emphasizes politeness in data analysis.
Functions like `please-plot`can be blocked and more polite versions such as
`may-I-please-have-a-plot`can be required using `options(politeness=99)`. We do
not recommend setting negative values for the politeness option should you use this package.]

```{r eval=FALSE}
## This file produces a plot relating the explanatory variable to the outcome.
thedata <- read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf("fig1.pdf") ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2 <- please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
```

Now, in the future if I wonder how 'that plot on page 10' was
created, I will know: (1) 'that plot' is from a file called
`fig1.pdf`and (2) `fig1.pdf`was created in
`fig1.R`. In a future where R still exists, changing the figure
will require quick edits of commands already written.  In a future
where R does not exist, I will at least be able to read the plain text
R commands and use them to write code in my new favorite statistical
computing language: R scripts are written in
[plain text](http://en.wikipedia.org/wiki/Plain_text), and plain
text is a format that will be around as long as computer programmers
write computer programs.^[Since R is open
  source, I will also be able to download an old version of R, download an
  old-fashioned open-source operating system (like Ubuntu 10), and
  run the old-fashioned statistical computing environment in the
  old-fashioned operating system in a virtual machine on my
  new-fashioned actual machine.]

Moreover, realize that *file names send messages* to your future
self. This means that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you will remove some regret from your future self. For example, if you are studying inequalit and protest, you might try naming a file something like  `inequality-and-protest-figures.R` instead of  `temp9`or
`supercalifragilisticexpialidocious`.  By the way, the extension
`.R` tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our
antique hard drives for files containing R scripts.

**Step 1** If we know the provenance of results,
future  or current collaborators can quickly and easily reproduce
and thus change and improve upon the work.

# No data analyst is an island for long.

Data analysis involves a long series of decisions. Each decision
requires justification.  Some decisions will be too small and
technical for inclusion in the published article itself. These need to
be documented in the code itself [@nagler1995coding]. Paragraphs
and citations in the publication will justify the most important
decisions. So, one must code to communicate with yourself and
others. There are two main ways to avoid forgetting the reasons you
did something with data: comment your code and tightly link your code
with your writing.^[One can also try the R command
  `put-it-in-my-mind(reason,importance="high")` to firmly place a
  reason for a decision into the mind of the analyst. I myself have
  not had much luck with this function.]

## Code is communication: Comment code

Comments --- unexecuted text inside of a script --- are a message to
collaborators (including your future self) and other consumers of your
work. In the above code chunk, I used comments to explain the lines to
readers unfamiliar with R and to remember that I had tried a different
specification but decided not to use it because adding the squared
term did not really change the substantive story arising from the
model.^[R considers text marked with `#` as a comment.]
Messages left for your future self (or near-future others) help
retrace and justify your decisions as the work moves from seminar
paper to conference paper to poster back to paper to dissertation and
onwards.

@@

Comment liberally. Comments are discarded when R runs analysis or LaTeX or pandoc
turns plain text into pdf or html files so only those who dig into
your work will see them.

## Code to communicate: Literate programming.

> Let us change our traditional attitude to the construction of
  programs: Instead of imagining that our main task is to instruct a
  computer what to do, let us concentrate rather on explaining to
  human beings what we want a computer to do.
  [@knuth1984literate, p.97]

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf file or some other format which displays page
images to ease reading rather than to encourage reanalysis and
rewriting. Eventually pieces of that report (tables, graphs,
paragraphs) ought to show up in, or at least inform, the publishable
paper. Re-creating those analyses by pointing, clicking, copying, or
pasting would invite typing error and waste time. Re-creating your
arguments justifying your analysis decisions would also waste
time.

More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the 'source code' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016 version of this paper, this idea is now widespread and made much easier than before via online services for code sharing and collaboration such as [GitHub](http://github.com) and [BitBucket](http://bitbucket.org).]

How might one avoid these problems?
[Literate programming](http://en.wikipedia.org/wiki/Literate_programming) is the practice of weaving code into a document ---
paragraphs, equations, and diagrams can explain the code, and the code
can produce numbers, figures, and tables (and diagrams and even equations and
paragraphs). Literate programming is not merely fancy commenting but
is about enabling the practice of programming itself to facilitate easy
reproduction and communication.

For example, in [@sec:data-analys-comp], I suggested that we know
where 'that plot on page 10 'comes from by making sure we had a
`fig1.pdf` file produced from a clearly commented plain text file
called something like `fig1.R`.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
\LaTeX~ markup with R code chunks to make things like
Figure~\ref{fig:giniprot}. This combination of LaTeX and R is called
Sweave [@leis:2005].^[Support for Sweave is included with R.]

```
This paper, for example, was written in plain text using Markdown
markup with R code chunks to make things like
Figure~\ref{fig:giniprot}. This combination of Markdown and R is
called [R markdown](http://rmarkdown.rstudio.com/).\footnote{Support for R markdown is
  included with R. For nicer documents, use the `rmarkdown` R package or click `Knit to PDF` or `Knit to HTML` using the free and open-source [RStudio IDE](http://rstudio.com) for R.}
\begin{figure}[h]
\centering
\`\`\`{r fig1plota,fig=TRUE}
##  Make a scatterplot of Protest by Inequality
with(good.df,plot(gini04,protest05,xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Mean Protest Activities\n(World Values Survey 2005--2010)'))
##  Label a few interesting points
with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(gini04,protest05,labels=Nation))
\`\`\`
\caption{Protest activity by income inequality [@norris2009data].}
\label{fig:giniprot}
\end{figure}
```

```{r fig1code,echo=FALSE,fig=FALSE}
##  Make a scatterplot of Protest by Inequality
with(good.df,plot(gini04,protest05,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
                  cex=.8))

##  Label a few interesting points
with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(gini04,protest05,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
```

\begin{figure}[H]
  \begin{center}
```{r fig1plot, ref.label="fig1code",echo=FALSE,fig=TRUE,width=4,height=4}
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
```
    \caption{Protest activity by income inequality @norris2009data.}
    \label{fig:giniprot}
  \end{center}
\end{figure}

By using `\label{fig:giniprot}`, I do not need to keep track of
the figure number, nor do extra work when I reorganize the document in
response to reviewer suggestions. Nor do I need a separate
`fig1.R`file or `fig1.pdf` file.  Tables and other
numerical results are also possible to generate within the source
code of a scholarly paper. Those who view the code for this essay will
see how Table~\ref{tab:protest} was also generated directly from a
regression object.\footnote{@beck2010reg inspired this
  particular presentation of a linear model.}

```{r maketable, echo=FALSE, results='hide'}
  lm1<-lm(protest05~gini04+meanpr,data=good.df)  #  Run the regression

@@

library(xtable)
thextab<-xtable(themat,align=c("l",rep(".",4)),digits=1,
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from [@norris2009data].",label="tab:protest")
```

```{r printtable,echo=FALSE,results='asis'}
  print(thextab,include.colnames=FALSE,
        hline.after=NULL,
        table.placement="H",
        add.to.row=list(
          pos=list(0,3),
          command=c(
            "\\toprule \n &
             \\multicolumn{1}{c}{Coef} &
             \\multicolumn{1}{c}{Std. Err.} &
             \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
            paste("& \\multicolumn{4}{c}{n: ",n,",",
                  "   resid.sd: ",signif(est.sigma),
                  ", R$^2$: ",r2,"} \\\\ \\bottomrule",sep="")
            )
          ))
```

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.^[The R project has a task view devoted to
  [reproducible research](http://cran.r-project.org/web/views/ReproducibleResearch.html) listing many of the different approaches to literate
  programming for R.}  If your workflow does not involve R, you can
still implement some of the principles here. Imagine creating a style
in Microsoft Word called 'code 'which hides your code when you print
your document, but which allows you to at least run each code chunk
piece by piece (or perhaps there are ways to extract all text of style
code from a Microsoft Word document). Or imagine just using some
other kind of indication linking paragraphs to specific places in code
files. There are many ways that creative people can program in a
literate way.

Literate programming need not go against the principle of modular data
analysis [@nagler1995coding]. In my own work I routinely have
several different Sweave files that fulfill different functions, some
of them create \LaTeX code that I can `\input` into my
`main.tex` file, others setup the data, run simulations, or allow
me to record my journeys down blind alleys. Of course, when we have
flying cars running on autopilot, perhaps something other than
Sweave will make our lives even easier. Then we'll change.

** Step 2 **

We analyze data in order to explain something about the world to other
scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis to human beings, we will do a better job
with the data analysis itself: we will learn as we focus on teaching,
and we will avoid errors and save time as we ensure that others
(including our future selves) can retrace our steps. A document than
can be 'run 'to reproduce all of the analyses also instills
confidence in readers and can more effectively spur discussion and
learning and cumulation of research.

# Meaningful code requires data.

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on ('load'/ 'open '/ 'get '/
'use') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
`load("thedata.rda")`or `read.csv("thedata.csv")`or some
equivalent line in them, and `thedata.csv`should be stored in
some place easy to find (like in the same directory as the file or
perhaps in `"Data/thedata.rda"`). Of course, it never hurts to
drop in a comment pointing to the data file.

Where should one store data files? An obvious solution is always to
make sure that the data file used by a command file is in the same
directory as the command file. More elegant solutions require all
co-authors to have the same directory structure so that
`load("Data/thedata.rda")` means the same thing on all computers
used to work on the project. This kind of solution is one of the
things that formal version control systems do well (as discussed
a bit more in [@sec:vers-contr-prev]).

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
[@nagler1995coding]. So, perhaps your analysis oriented files
will `load("cleandata.rda")`and a comment in the code will alert
the future you (among others) that `cleandata.rda`was created
from `create-cleandata.R` which in turn begins with
`read.csv(\url("http://www.greatfreedata.gov/dirtydata.csv"))`. Such
a data processing file will typically end with something like
`save("cleandata.rda")`so that we are doubly certain about the
provenance of the data.\footnote{Of course, if you need math or
  paragraphs to explain what is happening in these files, you might
  prefer to make them into Sweave files, for which the conventional
  extension is .Rnw. So you'd have `create-cleandata.Rnw` which
  might explain and explore the different coding decisions you made,
  perhaps involving a factor analysis and diagnostic plots.}

Now, if in the future we wonder where `cleandata.rda`came from, we
might search for occurrences of `cleandata` in the files on our
system. However, if such searching among files is a burden, an even
nicer solution is to maintain a file for each project called
`MANIFEST.txt` or INDEX.txt or README.txt which lists the
data and command files with brief descriptions of their functions and
relations.

** Step 3** We should know where the data came
from and what operations were performed on which set of data.

@@

```
DA NI=19 NO=199 MA=CM
LA=basic.lab
CM FI=basic.cov
```
